---
title: "R Notebook"
output: html_notebook
---

```{r}
# libraries 

library(tidyverse)
library(ggplot2)
library(lubridate)
library(tsibble)
library(stringr)
library(infer)
library(janitor)

```


```{r}
# read in data-sets 

residence_activity <- read_csv("raw_data/by_board_of_residence/inpatient_and_daycase_by_nhs_board_of_residence.csv") %>% clean_names()

treatment_activity <- read_csv("raw_data/by_board_of_treatment/inpatient_and_daycase_by_nhs_board_of_treatment.csv") %>% clean_names()

residence_age_sex <- read_csv("raw_data/by_board_of_residence/inpatient_and_daycase_by_nhs_board_of_residence_age_and_sex.csv") %>% clean_names()


residence_simd <- read_csv("raw_data/by_board_of_residence/inpatient_and_daycase_by_nhs_board_of_residence_and_simd.csv") %>% clean_names()


residence_speciality <- read_csv("raw_data/by_board_of_residence/inpatient_and_daycase_by_nhs_board_of_residence_and_specialty.csv") %>% clean_names()


beds <- read_csv("raw_data/by_board_of_treatment/beds_by_nhs_board_of_treatment_and_specialty.csv") %>% clean_names()

treatment_age_sex <- read_csv("raw_data/by_board_of_treatment/inpatient_and_daycase_by_nhs_board_of_treatment_age_and_sex.csv") %>% clean_names()

treatment_simd <- read_csv("raw_data/by_board_of_treatment/inpatient_and_daycase_by_nhs_board_of_treatment_and_simd.csv") %>% clean_names()


treatment_speciality <- read_csv("raw_data/by_board_of_treatment/inpatient_and_daycase_by_nhs_board_of_treatment_and_specialty.csv") %>% clean_names()


waiting_times <- read_csv("raw_data/by_board_of_treatment/monthly_ae_waitingtimes_202301.csv") %>% clean_names()




```

Questions 

The whole system: from what’s coming through the front door in A&E and then how it filters through; how many emergency admissions are there, how many electives, how does this impact the lengths of stay people have in hospital?

What does this mean for the number of people in intensive care and the number of beds?

Do people have longer lengths of stay from elective or emergency admissions?

```{r}

# for the app: 

# show A&E Attendances, Hospital Activity or Bed Occupancy 

# 2. for A&E attendances: 

        # 1. select: yearly or seasonal comparison 
        # 2. select: number of attendances or number meeting target. 
    
# 3. For Hospital Activity: 
   # select: whole period, years before pandemic, years during pandemic, seasons(years 2020-21), seasonal variation before pandemic.
   # select: type of patient: elective or emergency. 
   # select: number of episodes or length of stay: 
   # select: Intensive Care or Other departments
  

# For Bed Occupancy 

# select: whole period, years before pandemic, years during pandemic, or seasons (2020-21), seasons(before pandemic). 
# select: Intensive Care or Other departments
# select: total occupancy or percentage of all bed availability. 




```



```{r}

# clean monthly A&E Activity and Waiting times 

AE_activity_clean <-  
  waiting_times %>% 
  mutate(year = str_extract(month, "\\d{4}")) %>% 
  mutate(month = str_extract(month, "\\d{2}$")) %>% 
  mutate(month = case_when(month == "01" ~ "Jan", month == "02" ~ "Feb", month == "03" ~ "Mar", month == "04" ~ "Apr", month == "05" ~ "May", month == "06" ~ "June", month == "07" ~ "July", month == "08" ~ "Aug", month == "09" ~ "Sep", month == "10" ~ "Oct", month == "11" ~ "Nov", month == "12" ~ "Dec")) %>% 
  mutate(month = as.factor(month)) %>% 
  mutate(month = factor(month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "June", "July", "Aug", "Sep", "Oct", "Nov", "Dec"))) %>% 
  mutate(pandemic_years = case_when(year == "2020" ~ "Pandemic", year == "2021" ~ "Pandemic", year == "2017" ~ "Pre-pandemic", year == "2018" ~ "Pre-pandemic"))

```



```{r}

# clean treatment activity 

treatment_activity_clean <-
treatment_activity %>% 
  mutate(year = str_extract(quarter,"\\d{4}")) %>%
  mutate(quarter_extr = substr(quarter,5,6)) %>% 
  mutate(quarter_extr = case_when(quarter_extr == "Q1" ~ "Jan-Mar", quarter_extr == "Q2" ~ "Apr-June", quarter_extr == "Q3" ~ "July-Sep", quarter_extr == "Q4" ~ "Oct-Dec" )) %>% 
  mutate(quarter_extr = as.factor(quarter_extr)) %>% 
  mutate(quarter_extr = factor(quarter_extr, levels = c("Jan-Mar","Apr-June","July-Sep","Oct-Dec")))  %>% 
  mutate(pandemic_years = case_when(year == "2020" ~ "Pandemic", year == "2021" ~ "Pandemic", year == "2017" ~ "Pre-pandemic", year == "2018" ~ "Pre-pandemic"))
  
  

```

```{r}

# clean treatment specialty 

treatment_specialty_clean <-
treatment_speciality %>% 
  mutate(year = str_extract(quarter,"\\d{4}")) %>%
  mutate(quarter_extr = substr(quarter,5,6)) %>% 
  mutate(quarter_extr = case_when(quarter_extr == "Q1" ~ "Jan-Mar", quarter_extr == "Q2" ~ "Apr-June", quarter_extr == "Q3" ~ "July-Sep", quarter_extr == "Q4" ~ "Oct-Dec" )) %>% 
  mutate(quarter_extr = as.factor(quarter_extr)) %>% 
  mutate(quarter_extr = factor(quarter_extr, levels = c("Jan-Mar","Apr-June","July-Sep","Oct-Dec"))) %>% 
   mutate(pandemic_years = case_when(year == "2020" ~ "Pandemic", year == "2021" ~ "Pandemic", year == "2017" ~ "Pre-pandemic", year == "2018" ~ "Pre-pandemic"))



```

```{r}

# clean beds 

beds_clean <- 
  beds %>% 
  mutate(year = str_extract(quarter,"\\d{4}")) %>%
  mutate(quarter_extr = substr(quarter,5,6)) %>% 
  mutate(quarter_extr = case_when(quarter_extr == "Q1" ~ "Jan-Mar", quarter_extr == "Q2" ~ "Apr-June", quarter_extr == "Q3" ~ "July-Sep", quarter_extr == "Q4" ~ "Oct-Dec" )) %>% 
  mutate(quarter_extr = as.factor(quarter_extr)) %>% 
  mutate(quarter_extr = factor(quarter_extr, levels = c("Jan-Mar","Apr-June","July-Sep","Oct-Dec"))) %>% 
   mutate(pandemic_years = case_when(year == "2020" ~ "Pandemic", year == "2021" ~ "Pandemic", year == "2017" ~ "Pre-pandemic", year == "2018" ~ "Pre-pandemic"))
  

```

The A&E part 

```{r}

# PATIENTS COMING IN TO THE A&E 


# number of people coming in to A&E yearly since 2008

AE_activity_clean %>% 
  filter(!year %in% c("2023","2007")) %>% 
  group_by(year) %>% 
  summarise(yearly_count = sum(number_of_attendances_aggregate)) %>% 
  ggplot(aes(x = year, y = yearly_count))+
  geom_line(group =1)+
  theme_minimal()+
  labs(x = "", y = "")


```

```{r}

# number of people coming in to the A&E 2017 and 2018 compared to 2020 and 2021 . The number of people coming in to the A&E was lower before the pandemic. 


AE_activity_clean %>% 
  filter(pandemic_years %in% c("Pandemic", "Pre-pandemic")) %>% 
  ggplot(aes(x = pandemic_years, y = number_of_attendances_aggregate)) +
  geom_col()




```


```{r}

# seasonal variation before and during the pandemic 

AE_activity_clean %>% 
  filter(pandemic_years %in% c("Pandemic", "Pre-pandemic")) %>% 
  group_by(month, pandemic_years) %>% 
  summarise(sum_attendances = sum(number_of_attendances_aggregate)) %>% 
  ggplot(aes(x = month, y = sum_attendances, colour = pandemic_years, group = pandemic_years)) +
  geom_line()+
  theme_minimal()+
  labs(x = "", y = "", colour = "")



```
```{r}
# A&E targets met over years - pandemic has had a great cost it still hasn't recovered. 

AE_activity_clean %>% 
  filter(!year %in% c("2023","2007")) %>% 
  group_by(year) %>% 
  summarise(yearly_target_met = sum(number_meeting_target_aggregate)) %>% 
  ggplot(aes(x = year, y = yearly_target_met))+
  geom_line(group =1)+
  theme_minimal()+
  labs(x = "", y = "")

```
```{r}
# Targets met monthly comparison. Less patients meet target in the winters both in pandemic years and in pre-pandemic years. 

AE_activity_clean %>% 
  filter(pandemic_years %in% c("Pandemic", "Pre-pandemic")) %>% 
  group_by(month, pandemic_years) %>% 
  summarise(sum_target = sum(number_meeting_target_aggregate)) %>% 
  ggplot(aes(x = month, y = sum_target, colour = pandemic_years, group = pandemic_years)) +
  geom_point()+
  geom_line()+
  theme_minimal()+
  labs(x = "", y = "", colour = "")




```


```{r}


# people coming in to A&E monthly in pandemic years 


AE_activity_clean %>% 
  filter(year %in% c(2020,2021,2022)) %>% 
  group_by(month) %>% 
  summarise(monthly_count = sum(number_of_attendances_aggregate)) %>% 
  ggplot(aes(x = month, y = monthly_count))+
  geom_line(group =1)


#  A & E targets over years 

AE_activity_clean %>% 
  filter(!year == "2023") %>% 
  group_by(year) %>% 
  summarise(yearly_target_met = sum(number_meeting_target_aggregate)) %>% 
  ggplot(aes(x = year, y = yearly_target_met))+
  geom_line(group =1)


# monthly meeting target A & E during pandemic years 

AE_activity_clean %>% 
  filter(year %in% c(2020,2021,2022)) %>% 
  group_by(month) %>% 
  summarise(monthly_target_met = sum(number_meeting_target_aggregate)) %>% 
  ggplot(aes(x = month, y = monthly_target_met))+
  geom_line(group =1)


# Conclusion: for the A&E . less people came in over the winter. But less were also meeting target. Lack of staff availability or staff in the A&E occupied elsewhere? 

# INPATIENTS 

treatment_activity %>%  
  distinct(admission_type)


# Emergency number of patients over years 

treatment_activity_clean %>%
  filter(admission_type == "Emergency Inpatients", !year == 2017) %>% 
  group_by(year) %>% 
  summarise(admissions_year = sum(episodes)) %>% 
  ggplot(aes(x = year, y = admissions_year))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))

  

# Electives number of patients over years 

treatment_activity %>%
  filter(admission_type == "Elective Inpatients") %>% 
  group_by(quarter) %>% 
  summarise(admissions_quarter = sum(episodes)) %>% 
  ggplot(aes(x = quarter, y = admissions_quarter))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))

# CREATE GRAPH THAT SHOWS BOTH ELECTIVE AND EMERGENCY INPATIENTS COMING IN OVER TIME 

# number of episodes over hte the years 

treatment_activity_clean %>% 
  group_by(admission_type, year) %>% 
  summarise(n_episodes = mean(episodes)) %>% 
  filter(admission_type %in% c("Emergency Inpatients", "Elective Inpatients")) %>% 
  ggplot(aes(x = year, y = n_episodes, colour = admission_type, group = admission_type))+
  geom_point()+
  geom_line()+
  theme(axis.text.x = element_text(angle = 45))

# number of episodes over quarters

treatment_activity_clean %>% 
  group_by(admission_type, quarter_extr) %>% 
  summarise(n_episodes = mean(episodes)) %>% 
  filter(admission_type %in% c("Emergency Inpatients", "Elective Inpatients")) %>% 
  ggplot(aes(x = quarter_extr, y = n_episodes, colour = admission_type, group = admission_type))+
  geom_point()+
  geom_line()+
  theme(axis.text.x = element_text(angle = 45))


treatment_activity_clean %>% 
  distinct(year)

# length of stay over years 

treatment_activity_clean %>% 
  group_by(admission_type, year) %>% 
  summarise(mean_length = mean(length_of_stay)) %>% 
  filter(admission_type %in% c("Emergency Inpatients", "Elective Inpatients")) %>% 
  ggplot(aes(x = year, y = mean_length, colour = admission_type, group = admission_type))+
  geom_point()+
  geom_line()+
  theme(axis.text.x = element_text(angle = 45))


# length of stay over quarter 

treatment_activity_clean %>% 
  group_by(admission_type, quarter_extr) %>% 
  summarise(mean_length = mean(length_of_stay)) %>% 
  filter(admission_type %in% c("Emergency Inpatients", "Elective Inpatients")) %>% 
  ggplot(aes(x = quarter_extr, y = mean_length, colour = admission_type, group = admission_type))+
  geom_point()+
  geom_line()+
  theme(axis.text.x = element_text(angle = 45))


# Do people have longer lengths of stay from elective or emergency admissions?

# save filtered data 

treatment_activity_filt <- treatment_activity %>% 
  filter(admission_type %in% c("Emergency Inpatients", "Elective Inpatients")) %>% 
  mutate(admission_type = as.factor(admission_type))

# generate null distribution 

null_dist <- treatment_activity_filt %>% 
  specify(length_of_stay ~ admission_type) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 5000, type = "permute") %>% 
  calculate(stat = "diff in means", order= c("Emergency Inpatients", "Elective Inpatients"))

# calculate statistic 

observed_stat <- treatment_activity_filt %>% 
  specify(length_of_stay ~ admission_type) %>% 
  calculate(stat = "diff in means", order = c("Emergency Inpatients", "Elective Inpatients")) %>% 
  pull()

# observed_stat_p

# visualise 

null_dist %>% 
  visualise() +
  shade_p_value(obs_stat = observed_stat, direction = "two-sided")

# p-value 

p_value <- null_dist %>% 
  get_p_value(obs_stat = observed_stat, direction = "two-sided")

# p_value_p


# length of stay by average 

treatment_activity_filt %>% 
  group_by(admission_type, quarter) %>% 
  summarise(mean_mean_stay = mean(average_length_of_stay)) %>% 
  ggplot(aes(x = quarter, y = average_length_of_stay, colour = admission_type))+
  geom_point()+
  geom_line()+
  theme(axis.text.x = element_text(angle = 45))


```


```{r}

# What does this mean for the number of people in intensive care and the number of beds?

# number of people in intensive care -  number of people in intensive care drops 2019 Q4 to 2020 Q2

treatment_speciality %>% 
  filter(specialty_name == "Intensive Care Medicine") %>% 
  group_by(quarter) %>% 
  summarise(intensive_quarter = sum(episodes)) %>% 
  ggplot(aes(x = quarter, y = intensive_quarter))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))


# check for over years - number of people in intensive Care Medicine increases 


treatment_specialty_clean %>% 
  filter(specialty_name == "Intensive Care Medicine") %>% 
  group_by(year) %>% 
  summarise(intensive_year = sum(episodes)) %>% 
  ggplot(aes(x = year, y = intensive_year))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))


# check over quartas - more poeple in intensive care medicine in the winters of 2020 and 2021

treatment_specialty_clean %>% 
filter(specialty_name == "Intensive Care Medicine", year %in%c("2020", "2021")) %>% 
  group_by(quarter_extr) %>% 
  summarise(intensive_quarter_extr = sum(episodes)) %>% 
  ggplot(aes(x = quarter_extr, y = intensive_quarter_extr))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))


# number of beds percentage in intensive care 

 beds %>% 
  filter(specialty_name == "Intensive Care Medicine") %>% 
  group_by(quarter) %>% 
  summarise(beds_occ = mean(percentage_occupancy)) %>% 
  ggplot(aes(x = quarter, y = beds_occ))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))
 
 # beds over years 
 
 beds_clean %>% 
  filter(specialty_name == "Intensive Care Medicine") %>% 
  group_by(year) %>% 
  summarise(beds_occ = mean(percentage_occupancy)) %>% 
  ggplot(aes(x = year, y = beds_occ))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))
   
 
 # beds in intensive care over quortas in 2020 and 2021
 
  beds_clean %>% 
  filter(specialty_name == "Intensive Care Medicine", year %in% c(2020,2021)) %>% 
  group_by(quarter_extr) %>% 
  summarise(beds_occ = mean(percentage_occupancy)) %>% 
  ggplot(aes(x = quarter_extr, y = beds_occ))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))
 

# number bed occupancy average in intensive care - increases 2019 Q4 to 2020 Q2. "An occupied bed is an available staffed bed, which is either being used to accommodate an inpatient or reserved for a patient on pass". 


  beds %>% 
  filter(specialty_name == "Intensive Care Medicine") %>% 
  group_by(quarter) %>% 
  summarise(beds_occ = mean(average_occupied_beds)) %>% 
  ggplot(aes(x = quarter, y = beds_occ))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))
  
  # general bed occupancy pattern  by average
  
  beds %>% 
  group_by(quarter) %>% 
  summarise(beds_occ = mean(average_occupied_beds)) %>% 
  ggplot(aes(x = quarter, y = beds_occ))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))
  
  
  # general bed occupancy pattern by percentage 
  
  beds %>% 
  group_by(quarter) %>% 
  summarise(beds_occ = mean(percentage_occupancy)) %>% 
  ggplot(aes(x = quarter, y = beds_occ))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))
  
  beds_clean %>% 
    distinct(year)
  
#   Surgery Grouping, Women and Newborn Grouping, Intensive Care Medicine	, Mental Health Grouping	, Rehabilitation Medicine	
  
  
# beds seasons before pandemic . DO INTENSIVE CARE BEFORE AND AFTER THE PANDEMIC. 
  
  beds_clean %>% 
  filter(specialty_name == "Intensive Care Medicine", year %in% c(2019)) %>% 
  group_by(quarter_extr) %>% 
  summarise(beds_occ = mean(percentage_occupancy)) %>% 
  ggplot(aes(x = quarter_extr, y = beds_occ))+
  geom_point()+
  geom_line(group = 1)+
  theme(axis.text.x = element_text(angle = 45))
  
  treatment_specialty_clean %>% 
    filter(specialty_name == "Intensive Care Medicine") %>% 
    distinct(quarter)
  
```
```{r}

# Are there any specialties that have more stable activity patterns over time ?

treatment_speciality %>%
  group_by(specialty_name, quarter) %>%
  summarise(mean_length_episode = mean(length_of_episode)) %>%
  drop_na() %>%
  ggplot() +
  aes(
    x = quarter,
    y = mean_length_episode,
    colour = specialty_name, 
    group = specialty_name
  ) +
 geom_point()+
  geom_smooth(method = "lm", se = FALSE)+
  facet_wrap(~specialty_name)



```




```{r}

# emergency number of patients by sex 
treatment_age_sex %>% 
  filter(admission_type == "Emergency Inpatients") %>% 
  group_by(quarter, sex) %>% 
  summarise(admissions_quarter = sum(episodes)) %>% 
  ggplot(aes(x = quarter, y = admissions_quarter, colour = sex, group = sex))+
  geom_point()+
  geom_line()+
  theme(axis.text.x = element_text(angle = 45))

# length of stay by sex 

treatment_age_sex %>% 
  group_by(sex, admission_type, quarter) %>% 
  summarise(mean_length = mean(length_of_stay)) %>% 
  filter(admission_type %in% c("Emergency Inpatients", "Elective Inpatients")) %>% 
  ggplot(aes(x = quarter, y = mean_length, colour = sex, group = admission_type))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 45))


```

```{r}

# beds occupancy for other specialtites 

beds %>% 
  distinct(specialty_name)

 beds %>% 
  group_by(quarter, specialty_name) %>% 
  summarise(beds_occ = sum(average_occupied_beds)) %>% 
  filter(specialty_name %in% c("General Surgery", "Women and Newborn Grouping", "General Psychiatry (Mental Illness", "Mental Health Grouping", "Palliative Medicine", "Neurosurgery", "Cardiology","Respiratory Medicine", "Intensive Care Medicine")) %>% 
  ggplot(aes(x = quarter, y = beds_occ, colour = specialty_name, group = specialty_name))+
  geom_point()+
  geom_line()+
  theme(axis.text.x = element_text(angle = 45))

 
 ## look at mental health example 
 
 # bed occupancy average 
 
  beds_clean %>% 
  group_by(year, specialty_name) %>% 
  summarise(beds_occ = sum(average_occupied_beds)) %>% 
  filter(specialty_name == "Mental Health Grouping") %>% 
  ggplot(aes(x = year, y = beds_occ))+
  geom_point()+
  geom_line(group =1)+
  theme(axis.text.x = element_text(angle = 45))
  
  
  # bed occupancy precentages 
  
  beds_clean %>% 
  group_by(year, specialty_name) %>% 
  summarise(beds_occ = mean(percentage_occupancy)) %>% 
  filter(specialty_name == "Mental Health Grouping") %>% 
  ggplot(aes(x = year, y = beds_occ))+
  geom_point()+
  geom_line(group =1)+
  theme(axis.text.x = element_text(angle = 45))
 

```

```{r}

```







